Evaluating Informative Hypotheses With Equality and Inequality Constraint: The Bayes Factor Encompassing Prior Approach

### 2021-09-21 Documents/Bookdown/04-ext-model-comparison.Rmd
rearrange structure bookdown

---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Externalizing Problems {#ext}

In this chapter, we present the analyses to evaluate the different role of mother attachment and father attachment on children's externalizing problems. First, we discuss the appropriate models family to to take into account data characteristics. Next we conduct inference following three different approaches: 1) traditional Null Hypothesis Significance Testing (NHST) 2) Model Comparison using the AIC and BIC criteria 3) Bayes Factor with encompassing prior approach.

## Models Family Choiche {#model-choice}

Externalizing problems are computed as the sum of 10 items of the SDQ, obtaining discrete scores that range from 0 to 20. Thus, we should use appropriate discrete distribution such as the *Poisson* distribution or the *Negative Binomial*. In the Poisson distribution mean and variance are defined according to the same parameter $\lambda$. On the contrary, Negative Binomial has an extra parameter to adjust the variance allowing more flexibility. Considering data distribution (see Figure~\@ref(fig:plot-externalizing-dist)), we can observe that data have high dispersion with a long right tail. In this case, Poisson distribution would be a poor choice and we prefer Negative Binomial instead.

Again, considering data distribution (see Figure~\@ref(fig:plot-externalizing-dist)), we can observe an high peak of values at zero. Remember that this is not a clinical sample, thus it is expected that children majority has no problems or really few problem. We could question ourselves, however, whether a *Zero-Inflated* model may be appropriate

### Zero Inflated Negative Binomial

To evaluate the presence of zero inflation in our data, we compare the number of observed zeros and expected zeros in a Negative Binomial mixed effects model. We consider in the model the role of gender and the interaction between mother attachment and father attachment. Moreover, we consider the children's classroom ID as a random effect to account for teachers' different ability to evaluate children's problems. Using R formula syntax, we have
```{r echo = TRUE, eval=FALSE}
# model formula
externalizing_sum ~ gender + mother * father + (1|ID_class)
```

The model is fitted using the function `glmmTMB()` from the `glmmTMB` R-package [@brooksGlmmTMBBalancesSpeed2017]. Next, we compare the number of observed zero and expected zeros using an adapted version of the function `check_zeroinflation()` from the R-package `performance` [@ludeckePerformancePackageAssessment2021] that solves a small bug (see issue https://github.com/easystats/performance/issues/367).
```{r echo = TRUE, cache=TRUE}
my_check_zeroinflation(fit_ext_nb)
```

Results indicate that the model is slightly under-fitting the number of zeros. Now, we can try to fit a *Zero Inflated Negative Binomial* (ZINB) model and compare the performance of the two models. ZINB models are defined as
$$
y_{ij} \sim ZINB(p_{ij}, \mu_{ij}, \phi),
$$
where $p_{ij}$ is the probability of an observation $y_{ij}$ being an extra zero (i.e., a zero not coming from the Negative Binomial distribution) and $1-p_{ij}$ indicates the probability of a given observation $y_{ij}$ being generated form a Negative Binomial distribution with mean $\mu_{ij}$ and variance $\sigma_{ij}^2 = \mu_{ij} + \frac{\mu_{ij}^2}{\phi}$. Moreover, we have that
$$
p_{ij} = \text{logit}^{-1}(X_i^T\beta_p+ Z_j^Tu_p),\\
\mu_{ij} = \text{exp}(X_i^T\beta_{\mu}+ Z_j^Tu_{\mu}).
$$
That is, both $p$ and $\mu$ are modelled separately according to (possibly) different variables. In our case, we consider only the role of gender for $p$ (i.e., the probability of having externalizing problems depends on gender), whereas for $\mu$ the consider also the interaction between mother attachment and father attachment. In both cases, we consider the children's classroom ID as a random effect (teachers may differ in the ability to detect children's problems and quantify them). Using R formula syntax, we have
```{r echo = TRUE, eval=FALSE}
# formula for p
p ~ gender + (1|ID_class)

# formula for mu
mu ~ gender + mother * father + (1|ID_class)
```

The ZINB model is fitted using the function `zeroinfl()` from the `pscl` R-package [@zeileisRegressionModelsCount2008]. To compare the ZINB model and the Negative Binomial model we conduct an analysis of *Deviance*. Note that, in the case of generalized linear models (GLM), the deviance is the corresponding of the residual variance used in the traditional ANOVA in the case of linear models.
```{r echo=TRUE, cache=TRUE}
anova(fit_ext_nb, fit_ext_zinb)
```

Overall, results indicate that the ZINB model performs better than the Negative Binomial model. Thus, in the following analyses, we decide to use ZINB models.

## NHST

Following the traditional NHST approach, we consider the model previously defined that includes all effects of interest. That is the gender effect and the interaction between mother attachment and father attachment. Subsequently, we can run an analysis of deviance to evaluate the significance of the predictors using the function `Anova()` from the R-package `car` [@foxCompanionAppliedRegression2019].
```{r, echo =TRUE, cache=TRUE}
car::Anova(fit_ext_zinb)
```
Results indicate a statistically significant effect of gender and mother attachment. On the contrary, the interaction and father attachment are not significant. The model summary is reported below.
```{r, echo =TRUE, cache=TRUE}
summary(fit_ext_zinb)
```
To evaluate the effect of gender and mother attachment, the marginal predicted values according to gender and mother attachment are presented separately in Figure~\@ref(fig:plot-nhst-effects). Not that the marginal predicted values for gender are averaged over mother and father attachment effects. Whereas, the marginal predicted values for mother attachment are averaged over father attachment and gender effect.
```{r plot-nhst-effects, cache=TRUE, fig.asp=.65, message=FALSE, fig.cap="Marginal predicted values according to gender and mother attachment. Values are averaged over the other effects ($n_{subj} = 847$)."}
get_plot_zinb(model = fit_ext_zinb, attachment = "mother")
```

Post-hoc tests are run to evaluate differences between mother attachment styles. To do that we use the `contrast()` function from the `emmeans` R-package, considering pairwise comparisons and adjusting *p*-values according to multivariate *t*-distribution. This approach is less restrictive than the traditional *“Bonferroni”* method, as it determines the adjustment according to a multivariate *t*-distribution with the same covariance structure as the estimates. Results are reported below,
```{r, cache=TRUE, echo = TRUE}
emmeans::contrast(emmeans::emmeans(fit_ext_zinb, specs = ~ mother ),
                  "pairwise", adjust = "mvt")
```

Overall, results indicate that Males have more externalizing problems than Females and, regarding mother attachment, Fearful children have more problems than Secure children.

To evaluate the fit of the model to the data, we used $R^2$. In the case of generalized mixed-effects models, however, there are several definitions of $R^2$. We computed the *Marginal* $R^2$ and the *Conditional* $R^2$ as suggested by @nakagawaCoefficientDeterminationR22017. *Marginal* $R^2$ is concerned with the variance explained by fixed factors of the model, and *Conditional* $R^2$ is concerned with the variance explained by both fixed and random factors of the model. To do that we use the function `performance::r2()`.

```{r, echo=TRUE, message=FALSE, cache=TRUE}
performance::r2(fit_ext_zinb)
```

We can see that the actual variance explained by fixed effects is almost 10\%, not bad for psychology.

Considering attachment theoretical perspectives, results indicate only the role of mother attachment. Note, however, that traditional NHST does not allow us to evaluate evidence in favour of a hypothesis. Moreover, we actually have not tested our hypotheses but only the catch-all null hypothesis that *“nothing is going on”*.

## Model Comparison

@@

### Formalize Models

Following the same reasons as before (see Section~\@ref(model-choice)), we consider Zero Inflated Negative Binomial Mixed-Effects models. Again, we consider only the role of gender as a fixed effect and children's classroom ID as a random effect for $p$. Whereas, considering $\mu$, we define four different models to take into account the different theoretical perspectives:

@@

### AIC and BIC Results

@@

### Selected Model

@@

## Bayes Factor

To properly evaluate hypotheses with information regarding the expected direction of the effects, we use the Bayes Factor with encompassing prior approach. See the main article for a detailed introduction to this approach [TODO: add link article].

First, we define the encompassing model. Subsequently, we obtain the hypotheses matrices according to the informative hypotheses. Next, we compute the Bayes Factor and, finally, we describe the selected model.

### Encompassing Model

We define a Zero-Inflated Negative Binomial (ZINB) mixed-effects model to take into account the characteristics of the dependent variable and its distribution (see Section~\@ref(model-choice)).  Again, we consider only the role of gender as a fixed effect and children’s classroom ID as a random effect for $p$. Whereas, regarding $\mu$, we consider the interaction between mother and father attachment together with gender as fixed effects and children’s classroom ID as a random effect. In the R formula syntax, we have
```{r, echo=TRUE, eval=FALSE}
# formula for p
p ~ gender + (1|ID_class)

# formula for mu
mu ~ gender + mother * father + (1|ID_class)
```

#### Prior Choiche {-}

The prior choice is important for the parameters involved in the equality and inequality constraints. In our case, the parameters of interest (i.e., those related to mother and father attachment interaction) are unbounded. Thus, we can simply specify as prior a normal distribution with mean 0 and a given standard deviation. Considering the standard deviation, however, we have to choose a value so that the resulting prior is non-informative but without being excessively diffuse.

We can evaluate the consequences of different values choice considering the resulting prior predictions. To facilitate this step, we compute prior prediction considering only the intercept and a single parameter of interest. Remembering that the inverse link function (i.e., function that in a GLM transform the model linear prediction into the value on the original response scale) is the exponential function, we consider as intercept the value 1 because $exp(1) \approx 2.7$ that is close to the externalizing problems sample mean  `r ext_mean()`. In Table~\@ref(tab:table-prior-predict), summary information about prior predictions for different standard deviation values is reported.
```{r table-prior-predict}
get_table_prior_predict(format = params$format)
```

Considering that externalizing problems are bounded between 0 and 20, a reasonable prior is  $\mathcal{N}(0,3)$. as prior predicted values cover all possible values without including excessively large values. More diffuse priors would result in values with a higher order of magnitude and tighter priors would exclude plausible values. The influence of prior specification will be subsequently evaluated in a prior sensitivity analysis.

Regarding the other nuisance parameters (i.e., intercepts, random effects and shapes parameters) `brms` default priors are maintained. The resulting prior settings are
```{r prior-settings-encompassing-ext}
options(width = 300)
brms::prior_summary(encompassing_model_ext)
```

#### Posterior{-}
The encompassing model is estimated using 6 independent chains with 10,000 iterations (warm-up 2,000). To do that we use the `brm()` function from the `brms` R-package [@burknerBrmsPackageBayesian2017; @burknerAdvancedBayesianMultilevel2018a], which is based on STAN [@standevelopmentteamRStanInterfaceStan2020]. Summary of the encompassing model is presented below.
```{r}
summary(encompassing_model_ext)
```

### Hypothesis Matrices

For each informative hypothesis, we obtain a hypothesis matrix that translates equality and inequality constraints according to the encompassing model parametrization. Formalization of informative hypotheses and the procedure to derive hypothesis matrices are described in the main paper ([TODO: add link]).

Here we present the obtained hypothesis matrices were on the columns we have the parameters of the encompassing model (excluding the intercept) and each row express an equality constraint or an inequality constraint. Matrices row names follow this notation: names without square brackets indicate a constraint directly on the model parameter; names within square brackets indicate a group condition (that could be the resulting composition of more parameters). For example, `"M_Avoidant:F_Anxious"` indicates the actual interaction term of the model, whereas `"[M_Avoidant_F_Anxious]"` indicate the group condition. This is done because when assuming no interaction or no father attachment effect we set constraints directly on the model parameters, instead, when constraints involve group comparisons, we need to obtain the resulting conditions. Equality and inequality constraints are presented separately.

#### Null Hypothesis {-}

- **Equality matrix**, $R_{iE} = 0$ (each row is set to zero).
```{r cache=TRUE}
get_hypothesis_matrix("null", encompassing_model_ext)$eq
```

- **Inequality matrix**, $R_{iI} > 0$ (each row is set to grater than zero). There are no inequality constraints.

#### Monotropy Hypothesis {-}

- **Equality matrix**, $R_{iE} = 0$ (each row is set to zero).
```{r cache=TRUE}
get_hypothesis_matrix("monotropy", encompassing_model_ext)$eq
```

- **Inequality matrix**, $R_{iI} > 0$ (each row is set to grater than zero).
```{r cache=TRUE}
get_hypothesis_matrix("monotropy", encompassing_model_ext)$ineq
```

#### Hierarchy Hypothesis {-}

- **Equality matrix**, $R_{iE} = 0$ (each row is set to zero).
```{r cache=TRUE}
get_hypothesis_matrix("hierarchy", encompassing_model_ext)$eq
```

- **Inequality matrix**, $R_{iI} > 0$ (each row is set to grater than zero).
```{r cache=TRUE}
get_hypothesis_matrix("hierarchy", encompassing_model_ext)$ineq
```

#### Independence Hypothesis {-}

- **Equality matrix**, $R_{iE} = 0$ (each row is set to zero).
```{r cache=TRUE}
get_hypothesis_matrix("independence", encompassing_model_ext)$eq
```

- **Inequality matrix**, $R_{iI} > 0$ (each row is set to grater than zero).
```{r cache=TRUE}
get_hypothesis_matrix("independence", encompassing_model_ext)$ineq
```

#### Integration Hypothesis {-}

- **Equality matrix**, $R_{iE} = 0$ (each row is set to zero).
```{r cache=TRUE}
get_hypothesis_matrix("integration", encompassing_model_ext)$eq
```

- **Inequality matrix**, $R_{iI} > 0$ (each row is set to grater than zero).
```{r cache=TRUE}
get_hypothesis_matrix("integration", encompassing_model_ext)$ineq
```

### Centering and Adjusting

So far we have specified the encompassing prior, obtained the model posterior distribution, and defined the hypotheses matrices. Now, we need to transform our parameters of interest and center the distribution on the constraints focal points of interest. We apply the following transformation
$$
\beta = R\theta - r
$$
but we can ignore $r$ as in all our constraints it is always a vector of zeros.

Next, we get the adjusted prior and the posterior of the transformed parameters vector  $\beta$ (i.e., the parameters that identify the constraints) for each hypothesis. The adjusted prior is given by

$$
\pi_{adj}(\beta) \sim \mathcal{N}(0, \Sigma_{\beta}) = \mathcal{N}(0, R\Sigma_{\theta}R^T).
$$
Note that we set the mean vector to zero. The posterior is given by the same transformation
$$
Pr(\beta|Y) \sim \mathcal{N}(\hat{\beta}, \hat{\Sigma}_{\beta}) = \mathcal{N}(R\hat{\theta}-r, R\hat{\Sigma}_{\theta}R^T).
$$
See the main article for more details [TODO: add link].

This adjustment, however, requires the hypothesis matrix $R$ to be *full-row-rank* (i.e., all constraints are linearly independent). However, this is not the case of the Hierarchy Hypothesis. To overcome this issue, we follow the solution presented in the main article. First, define $R^*$ selecting the maximum number of independent rows. In this case, 15 contrast are independent

```{r cache=TRUE}
get_hypothesis_matrix("hierarchy", encompassing_model_ext)$hyp[1:15,]
```

```{r cache=TRUE}
R <- get_hypothesis_matrix("hierarchy", encompassing_model_ext)$hyp
```

The remaining contrasts, instead, are obtained as linear combinations of the other constraints. In particular,
```{r, echo = TRUE, cache = TRUE}
# Constraint 16: [M_Anx_F_Sec - M_Sec_F_Anx]
all(R["[M_Anx_F_Sec - M_Sec_F_Anx]",] == R["[M_Anx_F_Sec]", ] - R["[M_Sec_F_Anx]", ])

# Constraint 17: [M_Av_F_Sec - M_Sec_F_Av]
all(R["[M_Av_F_Sec - M_Sec_F_Av]",] == - R["[M_Anx_F_Sec - M_Av_F_Sec]", ] + R["[M_Sec_F_Anx - M_Sec_F_Av]", ] +  R["[M_Anx_F_Sec - M_Sec_F_Anx]", ])

# Constraint 18: [M_Fear_F_Sec - M_Sec_F_Fear] 
all(R["[M_Fear_F_Sec - M_Sec_F_Fear]",] == R["[M_Av_F_Sec - M_Sec_F_Av]", ] + R["[M_Fear_F_Sec - M_Av_F_Sec]", ] - R["[M_Sec_F_Fear - M_Sec_F_Av]", ])
```

Before computing the Bayes Factor, note that we have a set of comparable hypotheses as exists a common solution to the set of linear equations obtained by setting all hypothesis constraints equal to zero. The solution is the trivial solution of simply consider all parameters equal to zero. Finally, we do not need to standardize our parameters as they represent mean groups differences. See the main article for a detailed explanation [TODO: add link article].

### Results and Sensitivity

To compute the Bayes Factor we evaluate marginal densities and conditional probabilities as described in detail in the main article [TODO: add link article]. Bayes Factor and posterior probability of each hypothesis are reported in Table~\@ref(tab:table-bf-results-ext).

```{r table-bf-results-ext, cache = TRUE}
get_table_bf(bf_result = BF_weights_ext,
             path_img = "images/", problem = "ext",
             format = params$format)
```

Remember, however, that prior specification affects the Bayes Factor results. Therefore, we also evaluate the results considering different prior settings. In particular, we consider as possible priors for the parameters of interest:

- $\mathcal{N}(0,.5)$ - unreasonable tight prior
- $\mathcal{N}(0,1)$ - tighter prior
- $\mathcal{N}(0,3)$ - original prior
- $\mathcal{N}(0,5)$ - more diffuse prior
- $\mathcal{N}(0,10)$ - unreasonably diffuse prior

The results of the prior sensitivity analysis are reported in Table~\@ref(tab:table-sens-prior-analysis-ext).

```{r table-sens-prior-analysis-ext, cache = TRUE}
get_table_sens_analysis(summary_sensitivity_ext,
                        format = params$format,
                        bookdown = TRUE)
```

Overall results consistently indicate the Monotropy Hypothesis as the most supported by the data. However, we can observe two distinct patterns. As the prior gets more diffuse, the order of magnitude of the Bayes Factor comparing each hypothesis with the encompassing model increases. Moreover, the probability of the Null Hypothesis increases with more diffuse prior, whereas the probabilities of the Hierarchy, Independence and Integration Hypothesis increases with tighter priors.

To interpret these patterns, remember that order constraints are insensitive to the distribution specification as long as the distribution is symmetric and centred on the constraint focal point. On the contrary, equality constraints are highly affected by the prior definition (see the main article for more details[TODO: add link article]).

All the defined hypotheses include equality constraints. Thus, for more diffuse prior we observe that the order of magnitude of the Bayes Factor comparing each hypothesis with the encompassing model increases. Moreover, the hypothesis with a higher number of equality constraints (e.g., Null Hypothesis) will be favoured over hypotheses with a smaller number of equality constraints (e.g., Hierarchy, Independence and Integration Hypothesis).

### Selected Model

One of the limits of the Bayes Factor with the encompassing prior approach is that we only get the selected hypothesis but we do not obtain the actual estimates of the parameters posterior. To overcome this limit we rely on Bayesian inference that allows us to effectively estimate the model parameter posteriors.

This time in the model we consider only the role of gender and mother attachment as fixed effects of $\mu$. In the R formula syntax, we have
```{r, echo=TRUE, eval=FALSE}
# formula for p
p ~ gender + (1|ID_class)

# formula for mu
mu ~ gender + mother + (1|ID_class)
```
Again, we specify a normal distribution with mean 0 and standard deviation of 3, $\mathcal{N}(0,3)$, as prior for the beta parameters (i.e., those related to gender and mother attachment). Whereas, for the other nuisance parameters (i.e., intercepts, random effects and shapes parameters) `brms` default priors are maintained. The resulting prior settings are
```{r prior-settings-selcted-ext, cache=TRUE}
brms::prior_summary(brm_selected_ext)
```

The model is estimated using 6 independent chains with 6,000 iterations (warm-up 2,000). The model summary is presented below.
```{r cache=TRUE}
summary(brm_selected_ext)
```

Marginal effects are presented in Figure~\@ref(fig:plot-marginal-ext) and differences between mother attachment patterns are reported in Figure~\@ref(fig:plot-diff-ext).
```{r plot-marginal-ext, cache=TRUE, fig.asp=.5, out.width="95%", message=FALSE, fig.cap="Marginal predicted values according to gender and mother attachment ($n_{subj} = 847$)."}
plot_post_pred(post_pred_ext, problem = "Externalizing")
```

```{r plot-diff-ext, cache=TRUE, fig.asp=.5, message=FALSE, fig.cap="Predicted differences between mother attachment patterns ($n_{subj} = 847$)."}
plot_post_diff(post_pred_ext, problem = "Externalizing")
```

Overall, results indicate that Males have more externalizing problems than Females. Regarding mother attachment, Fearful, Avoidant, and Anxious children have more problems than Secure children. Moreover, Fearful children have more problems than Anxious children.

To evaluate the fit of the model to the data, we computed the *Bayesian* $R^2$ and using the function `brms::bayes_R2()`, and we present Posterior Predictions in Figure~\@ref(fig:plot-ppcheck-ext).

```{r, echo=TRUE, message=FALSE, cache=TRUE}
r2_ext
```

```{r plot-ppcheck-ext, cache=TRUE, fig.cap="Posterior predictive check ($n_{subj} = 847$).", message=FALSE,}
my_pp_check(brm_selected_ext, problem = "Externalizing")
```

We can see that the actual variance explained by fixed effects and random effect is around 15\%. Moreover, posterior predictive check indicates a good fit to the data.

#### Conclusions{-}

Considering attachment theoretical perspectives, results indicate only the role of mother attachment so we can support the Monotropy Theory.

## Conclusions Externalizing Problems

Overall, we obtained consistent results from the three different approaches. Males have more problems than Females and, regarding attachment, only mother attachment influence children's externalizing problems. In particular, we observed the following pattern: Secure children have the lowest level of problems, Anxious and Avoidant children have a similar, intermediate, level of problems, and Fearful children have the highest level of problems. Taken together, these results support the Monotropy Theory.

Although if the different approaches lead to apparently similar results, the rigorous interpretation of the results is very different.

- Considering the NHST approach, we actually only found that it is unlikely that mother attachment has no effect. Thus, we reject the null hypothesis but we can not quantify the evidence in favour of any of our hypotheses. 
- Only model comparison allow us to quantify the relative evidence of our models. Model comparison results clearly indicated evidence in favour of an effect of mother attachment but not father attachment. However, using information criteria we could not directly evaluate our informative hypotheses regarding the expected effects, but only the presence of any effect.
- Bayes Factor with encompassing prior approach allowed us to directly test our informative hypotheses regarding the expected effects. Results clearly selected the Monotropy theory as the most likely theory among those considered.

To summarize, the apparently identical results actually have a completely different meaning and what we learn from the data is very different. Hopefully, now it is clear that statistical inference is a complex process that requires careful thinking. In particular, to answer the questions we are actually interested in, we need to apply the appropriate statistical techniques.

>>>>

# Model Comparison {#model-comparison-ext}

@@

## Formalize Models

Following the same reasons as before (see Section~\@ref(model-choice-ext)), we consider Zero Inflated Negative Binomial Mixed-Effects models. Again, we consider only the role of gender as a fixed effect and children's classroom ID as a random effect for $p$. Whereas, considering $\mu$, we define four different models to take into account the different theoretical perspectives:

@@

## AIC and BIC Results

@@

## Selected Model

@@
